Trauma treatment delay

### 2023-12-15 manuscriptlongversion.Rmd
text

*Introduction:* Trauma remains a global health challenge, leading to
millions of deaths each year. The biggest impact can be seen in low to
middle income countries, but in high income countries, trauma systems
have been developed to lower the mortality and morbidity associated with
trauma. To further improve on these systems, conferences are held to
better the care patients receive and to identify errors made, called
opportunities for improvement, one of the more common being delayed
treatment. *Aim:* This retrospective cohort study aims to identify
patient level factors that are associated with receiving delayed
treatment among trauma patients.
<!-- The first sentence is a bit complicated. Try to simplify, break 
into multiple sentences --> *Material and Methods:* This study uses the
database of Karolinska university hospital, the equivalent of a level 1
trauma center, to run a logistical regression linking the outcome,
delayed treatment, to variables, observed patient level factors.
Patients treated for trauma at the hospital between 2012 and 2022 were
included, with ISS\>9 and/or trauma code activation. Exclusion criterias
were not being screened for OFI and missing values in any of the
variables, age, gender, ISS, SBP, RR, GCS and number of injuries.
*Results:* Higher ISS and RR was linked to higher rates of delay in
treatment (ISS, OR 1.03, 95% CI 1.01-1.06)(RR, OR 1.04, 95% CI
1.01-1.08), while a lower GCS also was linked to delayed treatment (OR
0.93, 95% CI 0.87-1.00). Number of injuries showed a correlation to
higher rates of delay, but not when adjusted for. The remaining
variables showed no correlation. *Conclusion:* Our study suggests that
it is mostly more severely injured and less awake patients that receive
delayed treatment.

*Inledning:* Trauma är fortfarande en global hälsoutmaning som skördar
miljontals liv varje år. Den största effekten kan ses i låg till
medelinkomstländer, men i höginkomstländer har traumasystem utvecklats
för att sänka morbiditeten och mortaliteten relaterat till trauma. För
att ytterliggare förbättra dessa system hålls konferenser för att
förbättra vården patienter får och identifiera misstag som gjorts, så
kallade förbättringsmöjligheter, där försenad vård är en av de
vanligaste. *Syfte:* Denna retrospektiva kohortstudie syftar till att
identifiera patientfaktorer kopplade till försenad vård bland
traumapatienter. *Material och Metod:* Denna studie använder Karolinska
universitetssjukhusets databas, motsvarande ett traumacenter på nivå 1,
för att köra en logistisk regression som kopplar utfallet, försenad
vård, till variabler, observerade patientnivåfaktorer. Patienter som
behandlades för trauma på sjukhuset mellan 2012 och 2022 inkluderades,
med ISS\>9 och/eller traumakodaktivering. Exklusionskriterier var om
patient inte screenades för ofi samt saknade värden i de använda
variablerna, ålder, kön, ISS, SBT, AF, GCS och antal skador. *Resultat:*
Högre ISS och AF var kopplade till högre grad av försenad vård (ISS, OR
1.03, 95% CI 1.01-1.06)(RR, OR 1.04, 95% CI 1.01-1.08), medan ett lägre
GCS också var kopplad till försenad vård (OR 0.93, 95% CI 0.87-1.00).
Antalet skador visade först ett samband med högre förseningsfrekvens,
men inte i den justerade analysen. De återstående variablerna visade
ingen korrelation. *Slutsats*: Vår studie tyder på att det mestadels är
svårare skadade och mindre vakna patienter som får försenad behandling.

@@

Trauma, defined as the clinical entity composed of physical injury and
the body's associated response [@gerdin_risk_2015], is and has long been
a major cause of death around the world [@WHO_2021].
<!-- The sentence below is long and overly complicated. Are you saying 
that trauma systems is an important part of healthcare and its implementation
has reduced mortality and morbidity?--> Within modern hospital care, the
trauma system stands as an important part in healthcare and is crucial
to lowering mortality and morbidity in injured patients through many
means, such as hospital care, patient follow up and prevention programs
[@Choi_2021].

@@

Today, trauma takes the lives of around 4,4 million people each year,
almost 8% of all deaths [@WHO_2021]. In the United States, trauma is the
4th leading cause of death among the general population and the leading
cause for people between the ages of 1 and 44 [@AAST_2020]. It is also
an important cause of hospitalization and morbidity among all age
groups, including seniors, and is responsible for an estimated 10% of
all years lived with a disability globally. This has a significant
burden on social and economic level, costing countries billions of US
dollars each year in healthcare and lost productivity [@connolly_2018].
Studies estimate the cost of trauma care being between 18,500\$ and
41,500\$ per patient in high income countries (HIC), depending on the
country [@Willenberg_2012]. However, it has been shown that this burden
is not evenly distributed between or within countries. Many social
factors such as age, sex and social status play a major role in the risk
of dying from trauma, with young men with low socio-economic status
being at most risk [@WHO_2021]. But it is not only patient level factors
that affect what effect trauma has on people. About 90% of all trauma
related deaths occur in low- and middle-income countries (LMIC), with
death rates by trauma also being higher than in HIC. Even within these
LMIC, people of poorer socio-economic status have higher death rates
from trauma. Problems identified within these countries that contribute
to these statisitics were infrastructure, education and training,
attributed to lack of funding, brain drain to HIC and lack of
availability of basic amenities [@Shanthakumar_2021]. In HIC where
funding and governance over the healthcare system is better designed,
functioning trauma systems and dedicated trauma centers exist that have
been shown to lower mortality but also improve functional outcome in
trauma patients [@Nirula_2006].

@@

<!-- Split long paragraphs -->

Trauma systems are infrastructures that exist to provide and optimize
care for injured patients starting with injury recognition and triage,
transport to appropriate trauma center, inpatient care and outpatient
follow-up. Beyond the clinical side, trauma systems work with outreach,
education and advocacy, data collection through registries, research,
funding, and disaster preparedness and response. A comprehensive and
functioning trauma system requires strong leadership and engagement at
the trauma center, regional and national level [@connolly_2018]. This
system is crucial to provide care for trauma patients, both in reducing
morbidity and mortality in this patient group. Earlier studies from
Sweden have shown that treating severely injured patients at a trauma
center is associated with a 41% lower adjusted 30-day mortality rate
compared to being treated at a non trauma care center due to them being
more capable of treating these patients, with potential survival benefit
increasing with higher injury severity [@Candefjord_2020]. Other studies
have shown similar results, showing that treating injured patients at
trauma centers is associated with a 15-30% decrease in mortality
[@Moran_2018; @Celso_2006; @MacKenzie_2006; @Haas_2012]. During the last
three decades, the introduction of trauma systems has contributed to
lowering the incidence of preventable death. This is attributed to the
improvements in care for acute brain injuries and bleeding control. The
incidence of late death because of sepsis and multiple organ failure has
also been lowered, possibly a result of better and earlier resuscitation
[@Asensio_2008]. Functioning trauma systems saves lives in the hospital,
but their work outside of the hospital is just as important. Data
collection from injured patients, such as mechanism of injury or
mortality, are essential for creating databases that can be used for
research. In turn, that research can be used for planning injury
prevention programs that target the most common injuries in the most
efficient ways, e.g. teen drivers, children, specific occupations etc.
These injury prevention programs can be planned on trauma center,
organizational or government level[@ACS].

@@

## Mortality and morbidity (MoM) conferences

<!-- Consider renaming this section as "Opportunities for improvement" -->

@@

```{=html}
<!-- Make sure the aim here is exactly the same as in the abstract. Also,
I suggest that your aim is to determine the associations between different 
patient level factors and receiving delayed treatment -->
```
The aim of this paper is to determine what different patient level
factors affect the risk of receiving delayed treatment at a trauma
center.

@@

Ethical permit is required and exists. The ethical review number is
2021-02541 and 2021-03531. An ethical problem that might be brought up
is the anonymity of the patients. In the registry, patients social
security number is present and could theoretically be used to identify
the patient in question. This will not be a problem though, as in the
analysis, the social security number is removed from the data before the
logistical regression is ran. Furthermore, the results will be presented
with patients grouped together, not individually. This will make it
impossible for anyone to identify patients. During the analysis, a VPN
was used to access the database to ensure that no data was leaked or
used for other purposes.

@@

The odds ratio is calculated both unadjusted and adjusted for other
variables<!-- What does "other" mean here?-->, as well as a 95%
confidence interval. The p value is also shown.

<!-- Use "associated" instead of correlated or linked etc-->

The analysis shows that some variables seem to have a correlation with
patients receiving delayed treatment. ISS and respiratory rate (RR) both
have a trend that patient with lower values seem to have a lower
incidence of delayed treatment, in both unadjusted (ISS, OR 1.04, 95% CI
1.02-1.05)(RR, OR 1.06, 95% CI 1.02-1.09) and adjusted analyses (ISS, OR
1.03, 95% CI 1.01-1.06)(RR, OR 1.04, 95% CI 1.01-1.08). GCS had the
opposite correlation, with higher GCS often having lower incidence of
delayed treatment, both unadjusted (OR 0.88, 95% CI 0.83-0.94) and
adjusted (OR 0.93, 95% CI 0.87-1.00). Number of injuries had the
correlation of higher number of injuries having higher incidence of
delayed treatment unadjusted (OR 1.06, 95% CI 1.02-1.10), but that
correlation was not seen when adjusted for. The three remaining
variables, age, gender and SBP, showed no correlation with the observed
outcome.

@@

These results show a possible correlation between some variables and the
presence of delayed treatment. A lower GCS had a correlation with
delayed treatment while a higher ISS had the same correlation. This
suggests that a more heavily injured and less awake patient is more
prone to receiving delayed treatment. Compared to a previous study
analysing the same data but with OFI as the outcome, this result is
similar to what they found using the same database [@Pat_factors]. As
the mentioned study suggests, this could be explained by that lower GCS
as well as higher ISS indicates severe trauma. More injured patients
often requires a more complex management and more interventions. More
interventions leave more room for delay to these interventions, why they
are more prone to delay in treatment. Respiratory rate also had the
correlation that lower respiratory rate correlated with lower rates of
delayed treatment among the patients. As the interval is \>10 to \<29 in
the triage guideline, both too low and too high respiratory rate is not
desirable[@EHCP2015]. This might have a similar explanation to ISS and
GCS, that patients with tachypnea might receive higher rates of delayed
treatment because they are more severely injured, therefore a higher
risk of missing a possible injury. The previous study also found this,
but the correlation dissapeared when adjusted for, which was
hypothesised being due to its shared variability with the ISS
variable[@Pat_factors]. In this study, this was the case for number of
injuries, as it and ISS is often correlated as a more severely injured
patient will both score higher on ISS and number of injuries. Therefore,
we hypothesize that the effect of number of injuries become
insignificant when adjusting for ISS.

@@

On discussing possible biases, misclassification bias might also be
<!-- Do not use contractions, i.e. can't, doesn't etc --> present in
this study, though it is something that we cannot control in this study.
As previously mentioned, the possible existence of an OFI is decided
during the MoM conferences, where they then add it to the database.
However, they might not always be right in their judgement and
misclassify from time to time, leading to patients landing in the wrong
group in the analysis. Also, some patients might have been missed even
before reaching the MoM conference. The patients included in the
conference are selected through audit filters and two specialized
nurses, and patients might be missed in this stage. This bias should not
be big enough to have a significant impact on the results, but can be
kept in mind.

```{=html}
<!-- Not so sure about this section. What do you mean that the logistic
regression assumes that the data is binomial? Do you mean that it requires 
a binary outcome? I think you discuss potential limitations well enough in 
the previous paragraphs, so you don't really need this -->
```
Another limitation with this study is that logistic regressions assumes
that the data is binomial, which the data used is not. The analysis
assumes a linear correlation between the dependent variable and the
independent variable, which is best analysed in a normal distribution.
However, Fischers test showed a significant difference between the
groups in 4 of the 7 variables, all of them being the ones who showed an
increased or decreased risk of delayed treatment, showing that these
results still are significant.

@@

@@

Discussing the generalisability of this study, we included the all the
trauma patients treated and Karolinska University Hospital and excluded
40.7% of that population based on that they had not been screened for
ofi. Additional patients were excluded because of missing data, but no
distinct groups except the not screened patients were removed. On that
basis, this population should be a general representation of trauma
patients at the equivalent of a level 1 trauma center, making this study
generalisable. This is in line with our aim, as it was to make it
applicable to the whole trauma patient population. However, as it is a
single center study, where its result only illustrates the situation in
Stockholm, it is less applicable nationally or globally. This does not
make it only applicable to the population of Stockholm, but is something
to keep in mind.

```{=html}
<!-- Think about what components of your study is generalisible and to 
where. I think that your results are generalisible to  a large extent, in 
the sense that severly injured patients are at higher risk of delayed 
treatment, also in other contexts -->
```
#### Future studies

>>>>

*Introduction:* Trauma remains a global health challenge, leading to
millions of deaths each year. The biggest impact can be seen in low to
middle income countries, but in high income countries, trauma systems
have been developed to lower the mortality and morbidity associated with
trauma. To further improve on these systems, conferences are held to
better the care patients receive and to identify errors made, called
opportunities for improvement, one of the more common being delayed
treatment. *Aim:* This retrospective cohort study aimed to determine the
associations between different patient level factors and receiving
delayed treatment. *Material and Methods:* This study used the database
of Karolinska university hospital to run a logistical regression linking
the outcome, delayed treatment, to variables, observed patient level
factors. Patients treated for trauma at the hospital between 2012 and
2022 were included, with ISS\>9 and/or trauma code activation. Exclusion
criterias were not being screened for OFI and missing values in any of
the observed variables. *Results:* Higher ISS and RR was linked to
higher rates of delay in treatment (ISS, OR 1.03, 95% CI 1.01-1.06)(RR,
OR 1.04, 95% CI 1.01-1.08), while a lower GCS also was linked to delayed
treatment (OR 0.93, 95% CI 0.87-1.00). Number of injuries showed a
correlation to higher rates of delay, but not when adjusted for. The
remaining variables showed no correlation. *Conclusion:* Our study
suggests that it is mostly more severely injured and less awake patients
that receive delayed treatment.

*Inledning:* Trauma är fortfarande en global hälsoutmaning som skördar
miljontals liv varje år. Den största effekten kan ses i låg till
medelinkomstländer, men i höginkomstländer har traumasystem utvecklats
för att sänka morbiditeten och mortaliteten relaterat till trauma. För
att ytterliggare förbättra dessa system hålls konferenser för att
förbättra vården patienter får och identifiera misstag som gjorts, så
kallade förbättringsmöjligheter, där försenad vård är en av de
vanligaste. *Syfte:* Denna retrospektiva kohortstudie syftade till att
identifiera associationen mellan patientfaktorer och försenad vård.
*Material och Metod:* Denna studie använde Karolinska
universitetssjukhusets databas för att köra en logistisk regression som
kopplar utfallet, försenad vård, till variabler, observerade
patientnivåfaktorer. Patienter som behandlades för trauma på sjukhuset
mellan 2012 och 2022 inkluderades, med ISS\>9 och/eller
traumakodaktivering. Exklusionskriterier var om patient inte screenades
för ofi samt om de saknade värden i de undersökta variablerna.
*Resultat:* Högre ISS och AF var kopplade till högre grad av försenad
vård (ISS, OR 1.03, 95% CI 1.01-1.06)(RR, OR 1.04, 95% CI 1.01-1.08),
medan ett lägre GCS också var kopplad till försenad vård (OR 0.93, 95%
CI 0.87-1.00). Antalet skador visade först ett samband med högre
förseningsfrekvens, men inte i den justerade analysen. De återstående
variablerna visade ingen korrelation. *Slutsats*: Vår studie tyder på
att det mestadels är svårare skadade och mindre vakna patienter som får
försenad behandling.

Keywords: Trauma, Delay to treatment, Patient factors.

@@

Trauma, defined as the clinical entity composed of physical injury and
the body's associated response [@gerdin_risk_2015], is and has long been
a major cause of death around the world [@WHO_2021]. The trauma system
plays an important part in modern healthcare and has shown to lower
mortality and morbidity in injured patients through many means, such as
hospital care, patient follow up and prevention programs [@Choi_2021].

@@

Today, trauma takes the lives of around 4,4 million people each year,
almost 8% of all deaths [@WHO_2021]. In the United States, trauma is the
4th leading cause of death among the general population and the leading
cause for people between the ages of 1 and 44 [@AAST_2020]. It is also
an important cause of hospitalization and morbidity among all age
groups, including seniors, and is responsible for an estimated 10% of
all years lived with a disability globally. This has a significant
burden on social and economic level, costing countries billions of US
dollars each year in healthcare and lost productivity [@connolly_2018].
Studies estimate the cost of trauma care being between 18,500\$ and
41,500\$ per patient in high income countries (HIC), depending on the
country [@Willenberg_2012].However, it has been shown that this burden
is not evenly distributed between or within countries.

Many social factors such as age, sex and social status play a major role
in the risk of dying from trauma, with young men with low socio-economic
status being at most risk [@WHO_2021]. But it is not only patient level
factors that affect what effect trauma has on people. About 90% of all
trauma related deaths occur in low- and middle-income countries (LMIC),
with death rates by trauma also being higher than in HIC. Even within
these LMIC, people of poorer socio-economic status have higher death
rates from trauma. Problems identified within these countries that
contribute to these statisitics were infrastructure, education and
training, attributed to lack of funding, brain drain to HIC and lack of
availability of basic amenities [@Shanthakumar_2021]. In HIC where
funding and governance over the healthcare system is better designed,
functioning trauma systems and dedicated trauma centers exist that have
been shown to lower mortality but also improve functional outcome in
trauma patients [@Nirula_2006].

@@

Trauma systems are infrastructures that exist to provide and optimize
care for injured patients starting with injury recognition and triage,
transport to appropriate trauma center, inpatient care and outpatient
follow-up. Beyond the clinical side, trauma systems work with outreach,
education and advocacy, data collection through registries, research,
funding, and disaster preparedness and response. A comprehensive and
functioning trauma system requires strong leadership and engagement at
the trauma center, regional and national level [@connolly_2018]. This
system is crucial to provide care for trauma patients, both in reducing
morbidity and mortality in this patient group. Earlier studies from
Sweden have shown that treating severely injured patients at a trauma
center is associated with a 41% lower adjusted 30-day mortality rate
compared to being treated at a non trauma care center due to them being
more capable of treating these patients, with potential survival benefit
increasing with higher injury severity [@Candefjord_2020]. Other studies
have shown similar results, showing that treating injured patients at
trauma centers is associated with a 15-30% decrease in mortality
[@Moran_2018; @Celso_2006; @MacKenzie_2006; @Haas_2012].

During the last three decades, the introduction of trauma systems has
contributed to lowering the incidence of preventable death. This is
attributed to the improvements in care for acute brain injuries and
bleeding control. The incidence of late death because of sepsis and
multiple organ failure has also been lowered, possibly a result of
better and earlier resuscitation [@Asensio_2008]. Functioning trauma
systems saves lives in the hospital, but their work outside of the
hospital is just as important. Data collection from injured patients,
such as mechanism of injury or mortality, are essential for creating
databases that can be used for research. In turn, that research can be
used for planning injury prevention programs that target the most common
injuries in the most efficient ways, e.g. teen drivers, children,
specific occupations etc. These injury prevention programs can be
planned on trauma center, organizational or government level[@ACS].

@@

## Opportunities for improvement

@@

This retrospective cohort study aimed to determine the associations
between different patient level factors and receiving delayed treatment.

@@

An ethical problem that might be brought up is the anonymity of the
patients. In the registry, patients social security number is present
and could theoretically be used to identify the patient in question.
This will not be a problem though, as in the analysis, the social
security number is removed from the data before the logistical
regression is ran. Furthermore, the results will be presented with
patients grouped together, not individually. This will make it
impossible for anyone to identify patients. During the analysis, a VPN
was used to access the database to ensure that no data was leaked or
used for other purposes. Ethical permit is required and exists. The
ethical review number is 2021-02541 and 2021-03531.

@@

The odds ratio is calculated both unadjusted and adjusted. A 95%
confidence interval was used with the p value shown in the table.

The analysis shows that some variables seem to have a association with
patients receiving delayed treatment. ISS and respiratory rate (RR) both
have a trend that patient with lower values seem to have a lower
incidence of delayed treatment, in both unadjusted (ISS, OR 1.04, 95% CI
1.02-1.05)(RR, OR 1.06, 95% CI 1.02-1.09) and adjusted analyses (ISS, OR
1.03, 95% CI 1.01-1.06)(RR, OR 1.04, 95% CI 1.01-1.08). GCS had the
opposite correlation, with higher GCS often having lower incidence of
delayed treatment, both unadjusted (OR 0.88, 95% CI 0.83-0.94) and
adjusted (OR 0.93, 95% CI 0.87-1.00). Number of injuries had the
association of higher number of injuries having higher incidence of
delayed treatment unadjusted (OR 1.06, 95% CI 1.02-1.10), but that
association was not seen when adjusted for. The three remaining
variables, age, gender and SBP, were not shown to be linked to the
outcome.

@@

These results show a possible correlation between some variables and the
presence of delayed treatment. A lower GCS had a correlation with
delayed treatment while a higher ISS had the same correlation. This
suggests that a more heavily injured and less conscious patient is more
prone to receiving delayed treatment. Compared to a previous study
analysing the same data but with OFI as the outcome, this result is
similar to what they found using the same database [@Pat_factors]. As
the mentioned study suggests, this could be explained by that lower GCS
as well as higher ISS indicates severe trauma. More injured patients
often requires a more complex management and more interventions. More
interventions leave more room for delay to these interventions, why they
are more prone to delay in treatment. Respiratory rate also had the
correlation that lower respiratory rate correlated with lower rates of
delayed treatment among the patients. As the interval is \>10 to \<29 in
the triage guideline, both too low and too high respiratory rate is not
desirable[@EHCP2015]. This might have a similar explanation to ISS and
GCS, that patients with tachypnea might receive higher rates of delayed
treatment because they are more severely injured, therefore a higher
risk of missing a possible injury. The previous study also found this,
but the correlation dissapeared when adjusted for, which was
hypothesised being due to its shared variability with the ISS
variable[@Pat_factors]. In this study, this was the case for number of
injuries, as it and ISS is often correlated as a more severely injured
patient will both score higher on ISS and number of injuries. Therefore,
we hypothesize that the effect of number of injuries become
insignificant when adjusting for ISS.

@@

On discussing possible biases, misclassification bias might also be
present in this study, though it is something that we cannot control in
this study. As previously mentioned, the possible existence of an OFI is
decided during the MoM conferences, where they then add it to the
database. However, they might not always be right in their judgement and
misclassify from time to time, leading to patients landing in the wrong
group in the analysis. Also, some patients might have been missed even
before reaching the MoM conference. The patients included in the
conference are selected through audit filters and two specialized
nurses, and patients might be missed in this stage. This bias should not
be big enough to have a significant impact on the results, but can be
kept in mind.

@@

Further analysing the results and it's clinical applications, the
results can be interpreted as that it is mostly injured patients with
affected consciousness that receive delayed treatment.

@@

Discussing the generalisability of this study, we included the all the
trauma patients treated and Karolinska University Hospital and excluded
40.7% of that population based on that they had not been screened for
ofi. Additional patients were excluded because of missing data, but no
distinct groups except the not screened patients were removed. On that
basis, this population should be a general representation of trauma
patients at the equivalent of a level 1 trauma center, making this study
generalisable. While the population only represents Karolinska
university hospital, the conclusion is should be applicable to the whole
trauma population.

#### Future studies
